COFFEE: an Optimizing Compiler for Finite Element Local Assembly

### 2014-06-25 TACO-paper/coffee.tex
Introduction revised

\section{Introduction}
In many fields, such as computational fluid dynamics, computational electromagnetics, and structural mechanics, phenomena are modelled by partial differential equations (PDEs). Numerical techniques, such as the finite volume method and finite element method, are widely employed to approximate solutions of these PDEs. Unstructured meshes are often used to discretize the equation domain, since their geometric flexibility allows solvers to be extremely effective. The solution is sought in each cell, or ``element'', of the discretized domain by applying suitable numerical operations. As the number of elements can be of the order of trillions \cite{Rossinelli2013}, a major issue is the time required to execute the computation, which can be hours or days. To address this problem, domain-specific languages (DSLs) have been developed. The successful port of Hydra, an industrial computational fluid dynamics application developed by Rolls Royce for turbomachinery design (based on finite volume method, roughly 50,000 lines of code and with meshes in excess of 100M edges), to OP2~\cite{op2-main}, demonstrates the effectiveness of the DSL approach in the implementation of PDE solvers~\cite{IstvanHydra}.

OP2 adopts a programming model in which computations are expressed through self-contained functions, referred to as ``kernels''. A kernel is applied to all elements in a set of mesh components, such as edges, vertices, or cells, with an implicit synchronization between the application of two consecutive kernels. On commodity multi-cores, a kernel is executed sequentially by a thread, while parallelism is achieved partitioning the mesh and assigning each partition to a thread. Similar programming and execution models are adopted in~\cite{pyop2isc},~\cite{Fenics},~\cite{fluidity_manual_v4},~\cite{lizst}. Kernel optimization is one of the major concerns in unstructured mesh applications. In this paper, we tackle this problem in the context of the finite element method.

We focus on improving the performance of local assembly (or ``assembly'', in the following), a fundamental step in the finite element method that covers an important fraction of the overall computation run-time, often in the range 30-60$\%$. During the assembly phase, the solution of the PDE is approximated by executing a suitable kernel over all elements in the discretized domain. In this work, we focus on relatively low order finite element methods, in which an assembly kernel's working set is usually small enough to fit the L1 cache. Low order methods are by no means exotic: they are employed in a wide variety of fields, including climate and ocean modeling, computational fluid dynamics, and structural mechanics. High order methods are often based upon special formulations of the finite element method, which result in the adoption of different sets of algorithms to solve PDEs~\cite{SPENCER}, and therefore are excluded from our study.

@@

Achieving high-performance is non-trivial. The complexity of the mathematical expressions, often characterized by a large number of operations on constants and small matrices, makes it hard to determine a single or specific sequence of transformations that is successfully applicable to all problems. The variation in loop trip counts and their typically small value further exacerbates the issue. A compiler-based approach is, therefore, the only reasonable option to obtain close-to-peak performance in a wide range of different local assembly kernels. Optimizations like generalized loop-invariant code motion, vector-register tiling, and expression splitting, as well as their composition, are essential, but none are supported by state-of-the-art polyhedral and vendor compilers. BLAS routines could theoretically be employed, although a fairly complicated control- and data-flow analysis is required to automate identification and extraction of matrix-matrix multiplies. In addition, as detailed in Section~\ref{sec:perf-eval-blas}, the small dimension of the  matrices involved and the potential loss in data locality can limit or eliminate the performance gain of this approach.

In order to overcome the constraints of the available compilers and specialized linear-algebra libraries, we have automated a set of generic and model-driven code transformations in COFFEE\footnote{COFFEE stands for COmpiler For FinitE Element local assembly.}, a compiler for optimizing local assembly kernels. COFFEE is integrated with Firedrake~\cite{firedrake-code}, a system for solving PDEs through the finite element method based on the PyOP2 abstraction~\cite{pyop2isc}. It supports all problems expressible with this framework. This allows us to evaluate our code transformations in a range of real-world problems, varying two key parameters that impact both solution accuracy and kernel cost: the polynomial order of the method (we investigate from $p=1$ to $p=4$) and the geometry of elements in the discretized domain (2D triangle, 3D tetrahedron, 3D prism).

Our experiments show that the original generated code for non-trivial assembly kernels is sub-optimal. Our cost-model-driven sequence of source-to-source code transformations, aimed at improving SIMD vectorization and register data locality, can result in performance improvements up to 1.5$\times$ over code in which only basic transformations have been performed, such as generalized loop-invariant code motion, padding, and data alignment, and up to 4.4$\times$ over original kernels. The contributions of this paper are
\begin{enumerate}
\item An optimization strategy for finite element local assembly. Our approach exploits domain knowledge and goes beyond the limits of both vendor and research compilers.
\item Design and implementation of a compiler that automates the proposed code transformations for any problems expressible in Firedrake.
\item Systematic analysis using a suite of examples of real-world importance, and evidence of significant performance improvements on two Intel architectures, a Sandy Bridge CPU and the Xeon Phi.
\end{enumerate}

@@

Despite the infinite variety of problem-specific assembly kernels which Firedrake can generate, it is still possible to identify common domain-specific traits that can be exploited for effective code transformations and SIMD vectorization. These include: 1) memory accesses along the three loop dimensions are always stride-1; 2) the \texttt{j} and \texttt{k} loops are interchangeable, whereas interchanges involving the $i$ loop require pre-computation of values (e.g. the $F$ values in Burgers) and introduction of temporary arrays, as explained in Section~\ref{sec:code-transf}; 3) depending on the problem being solved, the \texttt{j} and \texttt{k} loops could iterate over the same iteration space; 4) most of the sub-expressions on the right hand side of the element matrix computation depend on just two loops (either \texttt{i}-\texttt{j} or \texttt{i}-\texttt{k}). In Section~\ref{sec:code-transf} we show how to exploit these observations to define a set of systematic, composable optimizations.

@@

% Padding and Data Alignment
\subsection{Padding and Data Alignment}
The absence of stencils renders the element matrix computation easily auto-vectorizable by a vendor compiler. Nevertheless, auto-vectorization is not efficient if data are not aligned and if the length of the innermost loop is not a multiple of the vector length $vl$, especially when the loops are particularly small as in the case of assembly.

@@

We work around these limitations with source-level loop-invariant code motion. In particular, we pre-compute all values that an invariant sub-expression assumes along its fastest varying dimension. This is implemented by introducing a temporary array per invariant sub-expression and by adding a new loop to the nest. At the price of extra memory for storing temporaries, the gain is that lifted terms can be auto-vectorized as part of an inner loop. Given the short trip counts of our loops, it is important to achieve auto-vectorization of hoisted terms in order to minimize the percentage of scalar instructions, which could otherwise be significant. It is also worth noting that, in some problems (e.g. in Helmholtz), invariant sub-expressions along \texttt{j} are identical to those along \texttt{k}, and both loops iterate over the same iteration space. In these cases, we safely avoid redundant pre-computation since, as anticipated in Section~\ref{sec:background}, a property of our domain is that \texttt{j} and \texttt{k} loops share the same iteration space.

@@

\section{Related Work}
\label{sec:related-work}
The finite element method is extensively used to approximate solutions of PDEs. Well-known frameworks and applications include nek5000~\cite{nek5000-web-page}, the FEniCS project~\cite{Fenics}, Fluidity~\cite{fluidity_manual_v4}, and of course Firedrake. Numerical integration is usually employed to implement the local assembly phase. The recent introduction of DSLs to decouple the finite element specification from its underlying implementation facilitated, however, the development of novel approaches. Methods based on tensor contraction~\cite{FFC-Compiler} and symbolic manipulation~\cite{Francis} have been implemented. We have designed COFFEE to specifically target numerical integration because it has been demonstrated that it remains the optimal choice for a wide class of problems~\cite{quadrature1}.

>>>>

%Computational cost is a critical limitation in scientific computing, especially
%for finite element simulations. To provide one particular example we are
%particularly concerned about, it has been well established that mesh resolution
 Consequently, our aggressive optimization of
%local assembly, which may even take up to 60% of the overall FEM's execution
%time, directly impacts the performance of large-scale scientific simulations
%running on supercomputers.

\section{Introduction}
In many fields, such as computational fluid dynamics, computational electromagnetics, and structural mechanics, phenomena are modelled by partial differential equations (PDEs). Numerical techniques, like the finite volume method and the finite element method, are widely employed to approximate solutions of these PDEs. Unstructured meshes are often used to discretize the equation domain, since their geometric flexibility allows solvers to be extremely effective. The solution is sought by applying suitable numerical operations, or kernels, to the components of a mesh, such as edges, vertices, or cells. On standard clusters of multi-cores, typically, a kernel is executed sequentially by a thread, while parallelism is achieved partitioning the mesh and assigning each partition to a different node or thread. Such execution model, with minor variations, is adopted, for example, in \cite{pyop2isc}, \cite{Fenics}, \cite{fluidity_manual_v4}, \cite{lizst}.

%JUNK: , with synchronization handled at the level of the run-time support

The time required to apply the numerical kernels is a major issue, since the equation domain needs to be discretized into an extremely large number of elements, possibly order of trillions \cite{Rossinelli2013}, in order to obtain a satisfactory approximation of the PDE. For example, it has been well established that mesh resolution is critical in the accuracy of numerical weather forecasts. However, operational forecast centres have a strict time limit in which to produce a forecast (60 minutes in the case of the UK Met Office, who are partially funding this work), so producing efficient kernels has a direct scientific payoff in higher resolution, and therefore more accurate, forecasts. In finite elements, numerical kernels are executed during the local assembly phase (or ``assembly'', in the following), which can cover a significant fraction of the overall computation run-time, often in the range 30-60$\%$. In this paper, we aim at improving the performance of local assembly kernels in a wide variety of notably important PDEs though aggressive code transformation.

During the assembly phase, the solution of the PDE is approximated by executing a problem-specific kernel over all cells, or elements, in the discretized domain. In this work, we focus on relatively low order finite element methods, in which an assembly kernel's working set is usually small enough to fit the L1 cache. Low order methods are by no means exotic: they are employed in a wide variety of fields, including climate and ocean modeling, computational fluid dynamics, and structural mechanics. High order methods are often based upon special formulations of the finite element method, which result in the adoption of different sets of algorithms to solve PDEs~\cite{SPENCER}, and therefore are excluded from our study.

@@

Achieving high-performance is non-trivial. The complexity of the mathematical expressions, often characterized by a large number of operations on constants and small matrices, makes it hard to determine a single or specific sequence of transformations that is successfully applicable to all problems. The variation in loop trip counts and their typically small value further exacerbates the issue. A compiler-based approach is, therefore, the only reasonable option to obtain close-to-peak performance in a wide range of different local assembly kernels. Optimizations like padding, generalized loop-invariant code motion, vector-register tiling, and expression splitting, as well as their composition, are essential, but none are supported by state-of-the-art polyhedral and vendor compilers. BLAS routines could theoretically be employed, although a fairly complicated control- and data-flow analysis would be required to automate identification and extraction of matrix-matrix multiplies. In addition, as detailed in Section~\ref{sec:perf-eval-blas}, the small dimension of the  matrices involved and the potential loss in data locality can limit or eliminate the performance gain of this approach.

In order to overcome the constraints of the available compilers and specialized linear-algebra libraries, we have automated a set of generic and model-driven code transformations in COFFEE\footnote{COFFEE stands for COmpiler For FinitE Element local assembly.}, a compiler for optimizing local assembly kernels. COFFEE is integrated with Firedrake~\cite{firedrake-code}, a system for solving PDEs through the finite element method based on the PyOP2 abstraction~\cite{pyop2isc}. It supports all problems expressible with this framework, including equations that can be found at the core of real -world simulations, such as those used in our performance evaluation~\ref{sec:real-examples}. In evaluating our code transformations in a range of notable problems, we also vary two key parameters that impact solution accuracy and kernel cost: the polynomial order of the method (we investigate from $p=1$ to $p=4$) and the geometry of elements in the discretized domain (2D triangle, 3D tetrahedron, 3D prism).

%JUNK: 1.5$\times$ over code in which only basic transformations have been performed, such as generalized loop-invariant code motion, padding, and data alignment, and up to
Our experiments show that the original generated code for non-trivial assembly kernels is sub-optimal. Our cost-model-driven sequence of code transformations, aimed at improving SIMD vectorization and register data locality, can result in performance improvements up to 4.4$\times$ over original kernels. The contributions of this paper are
\begin{enumerate}
\item An optimization strategy for finite element local assembly. Our approach exploits domain knowledge and goes beyond the limits of both vendor and research compilers.
\item Design and implementation of a compiler that automates the proposed code transformations for any problems expressible in Firedrake.
\item Systematic analysis using a suite of examples of real-world importance, and evidence of significant performance improvements on two Intel architectures, a Sandy Bridge CPU and the Xeon Phi.
\end{enumerate}

@@

Despite the infinite variety of assembly kernels which Firedrake can generate, it is still possible to identify common domain-specific traits that can be exploited for effective code transformations and SIMD vectorization. These include: 1) memory accesses along the three loop dimensions are always stride-1; 2) the \texttt{j} and \texttt{k} loops are interchangeable, whereas interchanges involving the $i$ loop require pre-computation of values (e.g. the $F$ values in Burgers) and introduction of temporary arrays, as explained in Section~\ref{sec:code-transf}; 3) depending on the problem being solved, the \texttt{j} and \texttt{k} loops could iterate over the same iteration space; 4) most of the sub-expressions on the right hand side of the element matrix computation depend on just two loops (either \texttt{i}-\texttt{j} or \texttt{i}-\texttt{k}). In Section~\ref{sec:code-transf} we show how to exploit these observations to define a set of systematic, composable optimizations.

@@

% Padding and Data Alignment
\subsection{Padding and Data Alignment}
The absence of stencils renders the element matrix computation easily auto-vectorizable by a vendor compiler. Nevertheless, auto-vectorization is not efficient if data are not aligned and if the length of the innermost loop is not a multiple of the vector length $vl$, especially when the loops are small as in local assembly.

@@

We work around these limitations with source-level loop-invariant code motion. In particular, we pre-compute all values that an invariant sub-expression assumes along its fastest varying dimension. This is implemented by introducing a temporary array per invariant sub-expression and by adding a new loop to the nest. At the price of extra memory for storing temporaries, the gain is that lifted terms can be auto-vectorized as part of an inner loop. Given the short trip counts of our loops, it is important to achieve auto-vectorization of hoisted terms in order to minimize the percentage of scalar instructions, which could otherwise be significant. It is also worth noting that, in some problems (e.g. Helmholtz), invariant sub-expressions along \texttt{j} are identical to those along \texttt{k}, and both loops iterate over the same iteration space, as anticipated in Section~\ref{sec:background}. In these cases, we safely avoid redundant pre-computation.

@@

\section{Related Work}
\label{sec:related-work}
The finite element method is extensively used to approximate solutions of PDEs. Well-known frameworks and applications include nek5000~\cite{nek5000-web-page}, the FEniCS project~\cite{Fenics}, Fluidity~\cite{fluidity_manual_v4}, and of course Firedrake. Numerical integration is usually employed to implement the local assembly phase. The recent introduction of domain specific languages (DSLs) to decouple the finite element specification from its underlying implementation facilitated, however, the development of novel approaches. Methods based on tensor contraction~\cite{FFC-Compiler} and symbolic manipulation~\cite{Francis} have been implemented. We have designed COFFEE to specifically target numerical integration because it has been demonstrated that it remains the optimal choice for a wide class of problems~\cite{quadrature1}.
